---
title: "Ordered probit regression"
author: "Jake VanCampen"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: console
---


```{r setup, include=FALSE}
fig.dim <- 3
knitr::opts_chunk$set(fig.width=2*fig.dim,
                      fig.height=fig.dim,
                      fig.align='center',
                      message=FALSE)
```

## Introduction

This document will describe how to carry out *ordered probit regression* on a dataset of metric predictors and an ordinal predicted variable. This work is inspired from chapter 23 of [Doing Bayesian Data Analysis](https://sites.google.com/site/doingbayesiandataanalysis/about-the-author) (Kruschke JK). First I will describe the situation of ordinal predicted variable, then I will simulated the data with known paramters. A model will then be fit in Stan to estimate the known parameters, and the ability to accurately 'learn' these parameters will be assessed. An example of data that are described as ordinal are poll style ratings, but more generally are descrete ordered responses. 


I made up the following example to use for simulation of the data: 


A clinical study of 3000 melanoma tumors was carried out where each tumor was rated by a doctor on a severity scale of 1-10. Tumor biopsies were subject to RNA-seq, and For each tumor the normalized expression of a gene that is supposedly a predictor of phenotype was recorded where normalized expression is represented as read counts per kilobase per million (RPKM). In the model we are interested in the relationship between gene counts, and diagnotic severity.


To simulate these data we can look at the probability of ordinal predictors given by:

$$ p(y=k|\mu_k,\sigma,\{\phi_j\}) = \phi((\theta_k - \mu_k)/\sigma) - \phi((\theta_{k-1} - \mu_k)/\sigma) $$

Where the $\mu_k$ is described by a linear function of the predictor $x$ such that: 

$$ \mu_k = \beta_0 + \beta_1x $$

In the proposed example, $\mu$ will depend on the predictor $x$, the normalized gene counts.


We are interested in learning how the distribution of severity ratings vary with gene counts to see how gene counts are related to this severity phenotype. 


## Simulate the data

```{r}
library(LaplacesDemon)
set.seed(113)

# simulate counts for one gene
x <- rlnorm(3e3, 5, 0.5) 

# true parameters
beta_0 <- 0.5
beta_1 <- 0.02

# linear relationship of rating means to counts
mu_k <- beta_1 * x
  
# what's that look like
hist(mu_k)

# specify ordinal cutoffs (theta thresholds)
thetas <- c(1:10)
probs <- c(0.05,0.05,0.1,0.1,0.075,0.075,0.125,0.175,0.125,0.125)

y <- rep(thetas, probs*3e3)



thresh <- 1:10
sigma = 0.5

pr = matrix(rep(NA, 3e3*10), ncol=10)

for (i in 1:nrow(pr)) {
  pr[i, 1] = pnorm(thresh[1], mu_k[i], 1/sigma^2)
  for (k in 2:(length(thetas) -1)) {
    pr[i, k] = max(0, pnorm(thresh[k], mu_k[i], 1/sigma^2) - 
                      pnorm(thresh[k-1], mu_k[i], 1/sigma^2))
  }
  pr[i,length(thetas)] = 
    1 - pnorm(thresh[length(thetas)-1], mu_k[i], 1/sigma^2 ) 
}


dcat(pr, 1:10)

y <- rep(thetas, probs*3e3)

catmat <- matrix(c(thetas, probs), ncol=2)

dcat(c(1:10), catmat)

# calculate ratings from cutoffs

example <- read.table('../../stats2/data/OrdinalProbit.csv', sep=',',header = TRUE)

hist(example$X)
hist(example$Y)




### Make sure the metric predictors are standardized 
```



## Model the data in Stan

```{r}
library(rstan)
ord_mod <- "
data{
  int<lower=2> nlevs; // number of 'ratings'
  int<lower=0> N; // number of observations
  int<lower=1,upper=nlevs> y[N]; // the data
  vector[N] x; // gene counts metric predictor
}
parameters{
  real beta; // coefficeints
  ordered[nlevs-1] c; // categorical cutoffs
}
model{
  vector[nlevs] theta;
  for (n in 1:N) { 
    real eta;
    eta = x[n] * beta; // metric predictor
  
    // calculate using cum-normal Phi builtin
    theta[1] = 1 - Phi(eta - c[1]); // edge1

    for (k in 2:(nlevs-1))
      theta[k] = Phi(eta - c[k-1]) - Phi(eta - c[k]);

    theta[nlevs] = Phi(eta - c[nlevs-1]); // edge2
    y[n] ~ categorical(theta); 
  }
}
"

ord_prob_fit <- stan(model_code = ord_mod,
                     data=list(
                       nlevs=length(unique(example$Y)),
                       N = nrow(example),
                       m = length(unique(example$Y)),
                       y = example$Y,
                       x = example$X), 
                     chains = 3,
                     iter=1000)


summary(ord_prob_fit)

```












