---
title: "IA2"
author: "Jake VanCampen"
date: "`r format(Sys.Date(), '%B-%d-%Y')`"
output: html_document
---

```{r}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```


## Exploratory Data Analysis

A dataset concerned with thte effects of two fixed factors (genotype, and water) on the levels of three different proteins (ERD10, NCED9 and SRK2D) in Arabidopsis root tissues was explored to determine test the hypothesis that proteins in the root tissues are affected by genotype and water. 

First the data were read into R:

```{r}
library(tidyverse)
plant_data <- read_tsv('Plant_stress-1.tsv')
head(plant_data)
```



The two factors Genotype and Water should be cooerced to factors: 

```{r}
plant_data$Genotype <- factor(plant_data$Genotype)
plant_data$Water <- factor(plant_data$Water)
head(plant_data)
```


The effect of the two factors can be evaluated using boxplots for each protein in question:

### ERD10

```{r}
library(ggpubr)

wERD10 <- ggplot(plant_data)+
  geom_boxplot(aes(Genotype, log10(ERD10), fill = Genotype)) +
  ylab('ERD10 Content (Intensity)')+
  theme(legend.position = 'top')

gERD10 <- ggplot(plant_data)+
  geom_boxplot(aes(Water, log10(ERD10), fill = Water))+
  theme(legend.position = 'top', axis.title.y = element_blank())+
  scale_fill_brewer(palette = 'Accent')

iERD10 <- ggplot(plant_data)+
  geom_boxplot(aes(Genotype, log10(ERD10), fill = Water))+
  theme(legend.position = 'none')+
   scale_fill_brewer(palette = 'Accent')

inERD10 <- ggplot(plant_data)+
  geom_boxplot(aes(Water, log10(ERD10), fill = Genotype))+ 
  theme(legend.position = 'none', axis.title.y = element_blank())

ggarrange(wERD10,gERD10,iERD10,inERD10) %>% 
  annotate_figure(top = 'Effect of Genotype and Water on ERD10 content in Arabidopsis root tissue')

```


###  NCED9

```{r}
wNCED9 <- ggplot(plant_data)+
  geom_boxplot(aes(Genotype, log10(NCED9), fill = Genotype)) +
  ylab('NCED9 Content (Intensity)')+
  theme(legend.position = 'top')

gNCED9 <- ggplot(plant_data)+
  geom_boxplot(aes(Water, log10(NCED9), fill = Water))+
  theme(legend.position = 'top', axis.title.y = element_blank())+
  scale_fill_brewer(palette = 'Accent')

iNCED9 <- ggplot(plant_data)+
  geom_boxplot(aes(Genotype, log10(NCED9), fill = Water))+
  theme(legend.position = 'none')+
   scale_fill_brewer(palette = 'Accent')

inNCED9 <- ggplot(plant_data)+
  geom_boxplot(aes(Water, log10(NCED9), fill = Genotype))+ 
  theme(legend.position = 'none', axis.title.y = element_blank())

ggarrange(wNCED9,gNCED9,iNCED9,inNCED9) %>% 
  annotate_figure(top = 'Effect of Genotype and Water on ERD10 content in Arabidopsis root tissue')

```


### SRK2D

```{r}
wSRK2D <- ggplot(plant_data)+
  geom_boxplot(aes(Genotype, log10(SRK2D), fill = Genotype)) +
  ylab('SRK2D Content (Intensity)')+
  theme(legend.position = 'top')

gSRK2D <- ggplot(plant_data)+
  geom_boxplot(aes(Water, log10(SRK2D), fill = Water))+
  theme(legend.position = 'top', axis.title.y = element_blank())+
  scale_fill_brewer(palette = 'Accent')

iSRK2D <- ggplot(plant_data)+
  geom_boxplot(aes(Genotype, log10(SRK2D), fill = Water))+
  theme(legend.position = 'none')+
   scale_fill_brewer(palette = 'Accent')

inSRK2D <- ggplot(plant_data)+
  geom_boxplot(aes(Water, log10(SRK2D), fill = Genotype))+ 
  theme(legend.position = 'none', axis.title.y = element_blank())

ggarrange(wSRK2D,gSRK2D,iSRK2D,inSRK2D) %>% 
  annotate_figure(top = 'Effect of Genotype and Water on SRK2D content in Arabidopsis root tissue')

```



The two factors in this experiment (Genotype and Water) are both fixed factors. For the analysis effect of each of the factors on protein content, a Fixed-Factor Factorial ANOVA will be performed for each protein response variable.  It is clear form exploratory data analysis that there may be some interactions of the factors. 


There was some unhomogeneity of the variance between factor levels for SRK2D, and SRK2D. All data were thus log10-transformed for consistency. This transformation also improved normaility assumptions. 


For each protein response variable (ERD10, SRK2D and SRK2D), the null hypothesis that there is no effect of either fixed factor on the protein in question will be tested. 

## Two-Factor Factorial ANOVA 

### ERD10 

```{r}
# calculate the anova
ERD10_aov <- aov(ERD10 ~ Genotype + Water + Genotype * Water, plant_data)

# calculate anova table for each effect, and the interaction
anova(ERD10_aov)

# test the assumption of unequal variance
plot(ERD10_aov, which = 1) 

plant_data %>% 
  select(Genotype, Water, ERD10) %>% 
  group_by(Genotype, Water) %>% 
  summarise(mean = mean(ERD10),
            sem = (sd(ERD10))/sqrt((length(ERD10)))) -> ERD10_sum
  
# setbars
bars = aes(ymin=mean-sem, ymax=mean+sem)

# summarize with an interaction plot 
ggplot(data = ERD10_sum, aes(Genotype, mean, color = Water, group = Water))+
  stat_summary(fun.y = mean, geom = 'point') +
  stat_summary(fun.y = mean, geom = 'line') +
  geom_errorbar(bars, width = 0.2)+
  ylab('ERD10 Content (Intensity)') +
  ggtitle('Interaction plot of Water and Genotype on protein content')

```



There is a significant interaction of Water (F<sub>1,76</sub> = 6.8, *p* < 0.05), and the interaction of water and genotype (F<sub>1,76</sub> = 40.6, *p* < 0.001) on the content of ERD10 in Arabidopsis root tissues.



### SRK2D 

```{r}
# calculate the anova
SRK2D_aov <- aov(SRK2D ~ Genotype + Water + Genotype * Water, plant_data)

# calculate anova table for each effect, and the interaction
anova(SRK2D_aov)

# test the assumption of unequal variance
plot(SRK2D_aov, which = 1) 

plant_data %>% 
  select(Genotype, Water, SRK2D) %>% 
  group_by(Genotype, Water) %>% 
  summarise(mean = mean(SRK2D),
            sem = (sd(SRK2D))/sqrt((length(SRK2D)))) -> SRK2D_sum
  
# setbars
bars = aes(ymin=mean-sem, ymax=mean+sem)

# summarize with an interaction plot 
ggplot(data = SRK2D_sum, aes(Genotype, mean, color = Water, group = Water))+
  stat_summary(fun.y = mean, geom = 'point') +
  stat_summary(fun.y = mean, geom = 'line') +
  geom_errorbar(bars, width = 0.2)+
  ylab('SRK2D Content (Intensity)') +
  ggtitle('Interaction plot of Water and Genotype on protein content')

```

There is a significant effect of genotype (F<sub>1,76</sub> = 33.4, *p* < 0.001), water (F<sub>1,76</sub> = 40.5, *p* < 0.001), and the interaction of genotype and water (F<sub>1,76</sub> = 28.2, *p* < 0.001) on the content of SRK2D in Arabidopsis root tissues.

### NCED9

```{r}
# calculate the anova
NCED9_aov <- aov(NCED9 ~ Genotype + Water + Genotype * Water, plant_data)

# calculate anova table for each effect, and the interaction
anova(NCED9_aov)

# test the assumption of unequal variance
plot(NCED9_aov, which = 1) 

plant_data %>% 
  select(Genotype, Water, NCED9) %>% 
  group_by(Genotype, Water) %>% 
  summarise(mean = mean(NCED9),
            sem = (sd(NCED9))/sqrt((length(NCED9)))) -> NCED9_sum
  
# setbars
bars = aes(ymin=mean-sem, ymax=mean+sem)

# summarize with an interaction plot 
ggplot(data = NCED9_sum, aes(Genotype, mean, color = Water, group = Water))+
  stat_summary(fun.y = mean, geom = 'point') +
  stat_summary(fun.y = mean, geom = 'line') +
  geom_errorbar(bars, width = 0.2)+
  ylab('NCED9 Content (Intensity)') +
  ggtitle('Interaction plot of Water and Genotype on protein content')

```


There is a significant effect of genotype (F<sub>1,76</sub> = 159.6, *p* < 0.001), and water (F<sub>1,76</sub> = 6.7, *p* < 0.05) on the content of NCED9 in Arabidopsis root tissues; a mild, non-significant effect of the interaction of genotype and water was found (F<sub>1,76</sub> = 3.9, *p* < 0.1)


