---
title: "IA1"
author: "Jake VanCampen"
date: "`r format(Sys.Date(), '%b %d %Y')`"
output: pdf_document
---

```{r echo=FALSE}
knitr::opts_chunk$set(message=FALSE, warning = FALSE, include = FALSE, echo = FALSE)
```

```{r}
# import packages
library(tidyverse)
library(magrittr)
library(ggpubr)
library(knitr)
library(lmodel2)
library(broom)
zfish_raw <- read_tsv('zfish_diet_IA.tsv')
head(zfish_raw)

# clean the data
zfish_correct <- zfish_raw %>% 
  filter(., Weight < 4 & SL < 20)


# visualize with boxplots
bp_d <- ggplot(zfish_correct, aes(Diet, SL, fill = Diet))+ 
  geom_boxplot() + 
  ylab('Standard Length (mm)')+
  theme(
    axis.title.x = element_blank()
  )

bp_w <- ggplot(zfish_correct, aes(Diet, Weight, fill = Diet))+ 
  geom_boxplot()+ 
  ylab('Weight (mg)')+
  theme( 
    axis.title.x = element_blank()
    )

ggarrange(bp_d, bp_w, common.legend = TRUE, legend = 'right')  %>% 
          annotate_figure(., 
                          top = text_grob(
                          "Boxplots of zebrafish measurements by diet"), 
                          bottom = text_grob('Diet'))



# T tests for each 
sl_T <- t.test(zfish_correct$SL ~ zfish_correct$Diet)
wt_T <- t.test(zfish_correct$Weight ~ zfish_correct$Diet)


# format for reporting t test
t.report <- function(tt){
  tvalue <- tt$statistic %>% formatC(digits = 2, format = "f")
  pvalue <- tt$p.value %>% formatC(digits = 2, format = "E")
  if (round(tt$parameter, 0) == tt$parameter) {
    df <- tt$parameter
  } else {
    df <- formatC(tt$parameter, digits = 2, format = "f")
  }
  if (tt$p.value < 0.0005) {
    pvalue <- " < 0.001" 
  } else { 
    if (tt$p.value < 0.005) {
      pvalue <- paste0(" = ",
                       tt$p.value %>% 
                         formatC(., digits = 3, format = "f"))
    } else {
      pvalue <- paste0(" = ",
                       tt$p.value %>% 
                         formatC(., digits = 2, format = "f"))
    }
    } 
  paste0("*t*(",df,") = ",tvalue, ", *p*", pvalue)
}

```



## Results

This analysis explored the relationship between a protein enriched diet and mean fish size as measured by both standard length (mm) and weight (mg). The following hypotheses were tested here: 

$H{_0} \text{: Feeding zeebrafish an enriched diet will not significantly affect their mean size after two months}$
$H{_A} \text{: Feeding zebrafish an enriched diet will significantly affect their mean size after two months}$

Eploratory data analysis revealed two erroneous data points that were removed from the dataset. Zebra fish fed an enriched diet were found to have a significantly higher mean standard length (Welch two sample t-test, `r t.report(sl_T)`) and mean Weight (Welch two sample t-test, `r t.report(wt_T)`) compared to the non-enriched diet (Fig. 1, Table 1). The normality assumptions necessary to perform a parametric T-test were validated using a bootstrap resampling approach, and the null hypothesis that there is no significant effect of an enriched diet on mean fish size was rejected. 

There is a significant relationship between standard length (mm) and weight (mg) *F*<sub>1,196</sub> = 1643, *p* < 0.001, r-squared = 0.893 (Figure 2, Table 2). The relationship was fit using a type I linear model. In total these results accept the alternative hypothesis that zebrafish mean size is affected by a protein enriched diet as compared to the control, and show that both weight (mg) and standard length (mm) have a strong linear relationship showing the same increase in mean size with diet change.  


_______________________________________________________________________________


```{r include=TRUE} 
# visualize with boxplots
bp_d <- ggplot(zfish_correct, aes(Diet, SL, fill = Diet))+ 
  geom_boxplot() + 
  ylab('Standard Length (mm)')+
  theme(
    axis.title.x = element_blank()
  )

bp_w <- ggplot(zfish_correct, aes(Diet, Weight, fill = Diet))+ 
  geom_boxplot()+ 
  ylab('Weight (mg)')+
  theme( 
    axis.title.x = element_blank()
    )

ggarrange(bp_d, bp_w, common.legend = TRUE, legend = 'right')  %>% 
          annotate_figure(., 
                          top = text_grob(
                          "Boxplots of zebrafish measurements by diet"), 
                          bottom = text_grob('Diet'))


```

**Figure 1.** Boxplots showing the change in both standard length (mm) and Weight (mg) with a change in diet.






**Table 1.** Summary statistics for zebra fish measuremnets by diet.
```{r include=TRUE}
Summ_stats <- zfish_correct %>% 
  group_by(Diet) %>%
  summarise_at(vars(SL, Weight), c('mean', 'sd', 'var'))

kable(Summ_stats, 
      align = 'c', 
      digits = 3)

```







```{r include=TRUE}

#type I model fit 
zfish_lm <- lm(zfish_correct$Weight~zfish_correct$SL)
b <- format(data.frame(tidy(zfish_lm))$estimate[1],digit = 3) 
m <- format(data.frame(tidy(zfish_lm))$estimate[2],digits = 3)
r2 <- format(data.frame(glance(zfish_lm)$r.squared)[[1]], digits = 3)
eq <- paste("y=", m, "x +", b,',', 'r2', "=", r2)

stat_df <- tbl_df(data.frame(glance(zfish_lm)))
colnames(stat_df) <- c('r.squared','adj.r.squared','sigma','F-stat.','p.value','df','logLik','AIC','BIC','deviance', 'df.residual')


# plot the relationship between standard lenght and weight
corr <- ggplot(zfish_correct, aes(zfish_correct$Weight,
                          zfish_correct$SL,
                          color = Diet)) + 
  geom_point() + 
  geom_smooth(method = 'lm', color = 'black', size = 0.5) + 
  xlab('Standard Length (mm)') + 
  ylab('Weight (mg)')+
  annotate('text',x = 0.65, y = 6, label = eq)

sl_dens <- ggdensity(zfish_correct, 'SL', fill = 'Diet') +
           clean_theme()

wt_dens <- ggdensity(zfish_correct, 'Weight', fill = 'Diet') + 
           rotate() + clean_theme() 

# Arranging the plot
ggarrange(sl_dens, NULL, corr, wt_dens, 
          ncol = 2, nrow = 2,  align = "hv", 
          widths = c(2, 1), heights = c(1, 2),
          common.legend = TRUE, legend = 'bottom') %>% 
          annotate_figure(
            top = text_grob(
            'The relationship between weight and standard length over a change in diet',
            size = 14))
```


**Figure 2.** A type I linear model was fit to the relationship between standard length and weight, finding a significant positive relationship, *F*<sub>1,196</sub> = 1643, *p* < 0.001 (Table 2).


**Table 2.** Regression statistics for the type I linear model of the relationship between standard length (mm) and weight (mg) of the zebra fish. 
```{r include=TRUE}
kable(stat_df, align = 'c', digits = 3)
```



